Generalizing the Model View Controller Paradigm

### 2006-07-10 inst/paper/MVC.tex
Made changes to some of the figures and changed some of the text.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@18784 bc3139a8-67e5-0310-9ffc-ced21a209358

\abstract{ In this paper we consider a generalization of the model
  view controller (MVC) paradigm to allow for more flexible handling
  of submodels. Instead of using a single MVC we develop a tree
  structure of related MVCs, where each MVC has a single parent, but a
  parent can spawn many different child MVCs. Communication between
  parent and child MVCs allows for very flexible linking of the
  MVCs. We provide an implementation of our work in the R package
\Rpackage{MVC} available at \url{http://www.bioconductor.org}. }

@@

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data. Many authors
(citations) have suggested that linked interactive views of the data
provide a valuable and important set of tools for comprehending such
data. The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems. The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, easy extension to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset. In some cases the
natural data representation of the subset is in a form other than that
of the initial data set. This leads to complicatins in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a standard format. We found that adding new views under such
conditions could be quite problematic and often leads to very complex
code. Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree. Each parent can have multiple children, but the
children can have a single parent. Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow. Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that are linked in a natural way. Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the user, should they want it.

@@

Our development of this model was driven primarily by an interest in
analyzing high throughput biological data. Such data tend to be large
and very complex, and hence ideally suited to interactive
visualization. In order to provide the reader with some idea of the
sorts of interactions we would like to support we here provide a short
description of a type of experiment and some of the likely
interactions that would be of interest to a data analyst.

A gene expression experiment provides data on the expression of mRNA
abundance on thousands to tens of thousands of genes for some number
of samples. Usually there are many fewer samples than genes. There is
much that is known about the genes, such as what chromosome they are
located on, what sorts of activities they are engaged in and in many
cases we want to restrict attention to a specific subset of genes, or
to alter the view so that rather than being gene centric it becomes
centered on some biological aspect, such as chromosome location.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
more based on component software constructed within an object oriented
programming environment.

We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated
which we found results in a much shorter development time.

We wanted to provide a system that was usable and extensible by
relatively naive end users and hence have provided a graphical user
interface (GUI) which they can use to both carry out an analysis, but
also supports a variety of extensions, such as changing the actions
performed in response to events. We also wanted to support a more
sophisticate developer who might want to take our basic infrastructure
and create an interactive system, say for the interactive analysis of
time series data, using our infrastructure. Hence, all commands are
available both through the GUI and through command line functions.

@@

We first give our definitions of linked views, interactive views and
linked data sets are given. Our model is typically a statistical data
matrix which represents data on a number of variables, for each of a
number of samples.  By \textit{linked view} we mean that if a
component that represents data (such as a point on a plot or a row in
a spreadsheet) changes its appearance on one view, then the
corresponding component in all other views also changes appearance.
In our example, correspondence would be measured through the sample,
but there are other possibilities, as we shall explain later.  As an
example, suppose that the model consists of the gene expression levels
from say 50 samples and 10,000 genes.  The model can be represented as
a matrix of 50 columns and ten thousand rows.  We have two choices for
linking, either the entities can be the samples, or they can be the
genes. If the entities are the genes then when two scatter plots are
created as views, where on displays sample one versus sample two and
the second displays sample three versus sample four, then the points
on each plot that referred to the same gene are corresponding
components.  If instead we had chosen to link on samples, then we
could produce pairwise plots of gene expression values and would link
them through the sample IDs.

@@

By \textit{inked data sets} we mean that the data have more than one
conceptual grouping and that there is a relationship between these
different conceptual groupings.  Linked data sets are 
familiar to users of relational databases.  Data in a relational
database are typically stored in many linked tables with rows in
different tables linked to each other through a unique identifier
called a key.  In this example, each table can be considered a data set
and they are linked through the keys.  But the idea is more general
and we shall discuss other situations where the linked data sets do
not easily correspond to tables, however, we must have a map that goes
from the entities in one data set to those in the other as well as the
inverse map, in order to propagate signals between the MVCs.

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view. In the cartoon there are only two views, one of data set A and the
other of data set B, which is linked to data set A.  When a user
interacts with the scatter plot view of data set B, the response to
this interaction is for data set B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because data set B is linked
to data set A, data set B sends a message to data set A.  In Step 4,
the information from data set B is converted into information that
data set A understands.  In Step 5, this information is received by
data set A and data set A updates itself.  Finally, in Step 6, data
set 1 tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set B to update.  Then all views of data set B must be
      updated.  Next any data that are linked to data set B must be notified
      that data set B changed.  This information from data set B must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set A is notified that it should
      update in response to the change in data set B and then all views of
      data set A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newDatabaseTables.jpg}}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{newOneDatabaseTable.jpg}}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

@@

\section{Design Details An MVC}
\label{Sec:OneMVC}

@@

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that will be
needed by all models. Then, we created a number of subclasses, one for
each type of model. In Figure~\ref{Fig:Model} we show the inheritance
structure for the three types of models we have implemented, on based
on data frames, one based on graphs and one based on the expression
set class of data used by the Bioconductor project \citep{BIOC}.
%%FIXME: we need to send the complete paper2 to the BioC tech report
%%series and then cite it here, saying something like: Complete
%%implementation details are provided in XXX.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{ModelClass.jpg}}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views and all
must respond to a certain subset of events so we have implemented this
class by first creating a single, virtual class which the specialized
views extend.  As for our MVC classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{newViewClass.jpg}}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

As for specific view classes, the \Rclass{spreadView} class represents
a spreadsheet view of the data.  This view only makes sense for models
that are a two dimensional data structure, such as a matrix or a data
frame. Other views we have implemented include: scatterplot
(\Rclass{sPlotView}), graph (\Rclass{graphView}), and heatmaps
(\Rclass{heatmapView}).

Since all views are expected to be interactive they have methods that
correspond to the different supported user interactions.  The events
we are most interested in responding to include key press events,
where a certain configuration of keys are pressed; button click
events, where one of the mouse buttons is clicked over the view; and
mouse movement events, where the cursor moves over the view.  When an
event occurs it creates a signal, that is caught by the signal
handler.  The signal handler then calls the callback function and it
is in the callback function that the corresponding method is called.
For example, when the user clicks the mouse button and the cursor is
over a scatter plot view, it is the scatter plot's
\Rfunction{clickEvent} method that is invoked.  Using the information
that is passed to it, the method converts the pixel coordinates into
user coordinates and then finds the closest object in the plot.

All views need an update method so they can alter themselves when the
model sends them a message that the data have been altered. The
actions taken depend on the view, for example for a spread sheet view
the method may select or unselect a row on the spreadsheeet while for
a scatterplot view the method may change the appearance of a point on
the plot.  The inheritance structure of the view classes is
extensible.  For instance, if a user wanted to add a new plot view,
say a density plot, they can create a new view class that inherits
from \Rclass{plotView}, perhaps called \Rclass{dPlotView}.

@@

In our implementation we provide a control window, for users that want
to interact via the GUI.  It has menu items corresponding to
\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
\texttt{Window}.  These menus always appear on the control window, but
the items under each menu may change depending on the type of model
that is active.  For example, the \texttt{Display} menu allows users
to create views of the active model, but only certain views are
appropriate for a type of model, as mentioned in
Section~\ref{Ssec:OneViews}.  If the active model was a data frame
model (\Rclass{dfModel}), then the views that the user could create
are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
make sense.  Thus, when the model in the ``active'' \Robject{MVC} is
of class \Rclass{dfModel}, then the Display menu has menu items to
create a scatter plot and a spreadsheet.  Also, if the ``active''
\Robject{MVC} is of class \Rclass{dfModel}, then the command line
functions only allow the user create a scatterplot or a spreadsheet
view of the active model, since these are the only ones we have
implemented.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
In our design a callback function takes only one parameter. Further,
within a specific MVC all callback functions must have the same
signature so that different callback functions can be interchanged to
allow different responses to an event.  The class of that parameter
depends on the type of model represented by the \Robject{MVC}.  For
example, for a graph model, the parameter corresponds to a node in the
graph, while for a data frame model, the parameter is of class
\Rclass{character}, representing a case.

%%FIXME: I thought we talked about this, and found it to be a little
%%problematic, as the way this sounds every click will ask me to set a
%%color. Somehow, some of these are essentially global options that
%%need to be set once, and then used.
Any other information a callback function may need is set using the
preprocessing functions.  The preprocessing functions take no
parameters.  The purpose of the preprocessing functions is to set
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions and if no preprocessing
functions are needed, then this slot is NULL.  If the callback
function sets the color of a node in a graph, then a preprocessing
function is needed to determine what color is used.  So in this
example, the preprocessing function opens a color browser to allow the
user to choose what the new color is.

%%FIXME: any chance you could elucidate here - what events get
%%excluded, and why are we wanting to be the "event police?"
Users can link actions to events either through the GUI or via
commandline functions.  Note that not all callback functions make
sense for each event so the controller environment also stores a list,
for each even,t of appropriate instances that can be linked to that
event.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{fullMVCClass.jpg}}
    \caption{ The MVC Class. }
    \label{Fig:MVCClass}
  \end{center}
\end{figure}

%%FIXME: at this point we could perhaps even include a real example
%%with pictures of what is described below?
Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects are related to each other through a
parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
picture, `model1' is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
`model2.1' and `model2.2'. Then from `model2.1', three
\Robject{MVC} objects were created, `model3.1', `model3.2', and
`model3.3'.  As an example, model1 could be gene 
expression data, model2.1 could be gene expression data that is a subset of
model1 (based on filtering), model2.2 could be derived by performing
multidimensional scaling (MDS) on model1, model 3.1 could be the gene ontology
(GO) graph representing the molecular functions of the genes in model2.1,
model 3.2 could be the KEGG pathways that the genes in model2.1 are involved
in, and model 3.3 could be the chromosome locations of the genes in model2.1.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{Hierarchy2.jpg}}
    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
    \label{Fig:Hier}
  \end{center}
\end{figure}

@@

%%FIXME: this needs a few more words. What do these messages do, and
%%why are they the only ones we care about?
Currently, there are two types of messages within an MVC: an add
message and an update message.  Both messages request that some
modification be carried out.  Since adding a view requires different
information and methods from adding a model (and similarly with
updating views versus models), it makes sense to have two separate
message classes for these elements.  These messages can be sent to
either a model, or a view and their effect is quite different.

>>>>

\abstract{ In this paper we consider a generalization of the model
  view controller (MVC) paradigm to allow for more flexible handling
  of submodels. Instead of using a single MVC we develop a tree
  structure of related MVCs, where each MVC has a single parent, but a
  parent can spawn many different child MVCs. Communication between
  parent and child MVCs allows for very flexible linking of the
  MVCs. We provide an implementation of our work in the R package
\Rpackage{MVCClass} available at \url{http://www.bioconductor.org}. }

@@

There are many ways to visualize multivariate data when more than
two or three dimensions are needed to view the data. Many authors
(citations) have suggested that linked, interactive views of the data
provide a valuable and important set of tools for comprehending such
data. The model view controller (MVC) paradigm is now a standard
mechanism for implementing such systems. The paradigm has proven very
effective and allows for clear separation of different components.
In this paper we consider an extension of the MVC paradigm to allow
for, what we believe to be, an easy expansion to linked data sets.

The sorts of problems we are interested in, are those where a user
might begin with a large data set, identify some interesting subset
and want to create interactive views of that subset. In some cases the
natural data representation of the subset is in a form other than that
of the initial data set. This leads to complications in the usual MVC
paradigm, where the data (i.e. the model) are often presumed to be in
a standard format. We found that adding new views under such
conditions could be quite problematic and often leads to very complex
code. Hence we considered an approach that created new instances of
the MVC that were linked to the parent that spawned them.

In particular, our implementation consists of a hierarchy of MVCs,
stored as a tree. Each parent can have multiple children, but the
children can have a single parent. Communication between parent and
child is bidirectional, and a variety of different filters can be used
to control flow. Thus, we can devise a number of purpose built MVC
classes, each with a model appropriate to the specific data of
interest, that are linked in a natural way. Interacting with the view
in a child MVC, first sends a message to the child MVC, which will
update all of its views, but the message will also propagate to the
parent MVC (and further up to the root). The parent MVC can choose to
react, or not, to messages from its children -- all control being in
the hands of the users, should they want it.

@@

Our development of this model was driven primarily by an interest in
analyzing high throughput biological data. Such data tend to be large
and very complex, and hence ideally suited to interactive
visualization. In order to provide the reader with some idea of the
sorts of interactions we would like to support, we provide here a short
description of a type of experiment and some of the likely
interactions that would be of interest to a data analyst.

A gene expression experiment provides data on the expression of mRNA
abundance on thousands to tens of thousands of genes for some number
of samples. Usually there are many fewer samples than genes. There is
much that is known about the genes, such as what chromosome they are
located on and what sorts of activities they are engaged in, and in many
cases we want to restrict attention to a specific subset of genes, or
to alter the view so that rather than being gene centric it becomes
centered on some biological aspect, such as chromosome location.

@@

%% FIXME: perhaps we could have a heatmap and a few of the other views
%% here?
%% EAW: put them below
There is often substantial interest in relating gene expression data
to such a graph based model, or to chromosomal locations, or to
activity in known biological pathways. Data mappings for this latter
aspect can be obtained from different sources such as \cite{KEGG}.
Further, users typically want to begin with a heatmap view. Then from
that derive a GO graph view, and additionally a chromosomal location
view. In order to interact with the GO graph, the primary unit is the
node in that graph (and sometimes one expects with the edges), while
for the heatmap view the basic element is generally the gene.  In Figures
\ref{Fig:heatmap} and \ref{Fig:GOgraph}, a heatmap view of gene expression
from Chronic Lymphocytic Leukemia (CLL) patients and the corresponding GO
graph are shown, respectively.  By interacting with the heatmap to color one
gene blue (shown as gene PDE8A on the heatmap), which causes the corresponding
GO nodes to be colored blue, one can easily see how this gene relates to the
GO terms.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=5in]{heatmap.jpg}
    \caption{ Heatmap view of gene expression from Chronic Lymphocytic
      Leukemia (CLL) patients. }
    \label{Fig:heatmap}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4.3in, width=4.3in]{GOgraph.jpg}
    \caption{ Gene Ontology (GO) graph that is associated with the genes
      displayed in the heatmap.  Notice that one gene (PDE8A) is colored blue
      in the heatmap and all of the corresponding GO terms are colored blue in
      the graph. }
    \label{Fig:GOgraph}
  \end{center}
\end{figure}

@@

We found that constructing a single MVC that would encompass all
anticipated views was very cumbersome. It would also be inefficient
since a MVC model for the GO graph, or for a chromosome view would
also be of substantial interest, regardless of whether there was an
initial heatmap view or not. Thus, we looked for a solution that was
based more on component software constructed within an object oriented
programming environment.

We had several objectives, or design criteria. They were that the
software should be extensible in a high level language. We chose R as
the language for our implementation and wanted to allow users that
knew no other languages to be able to extend and enhance the
system. Therefore, we built our software on top the \Rpackage{RGtk} package of
D. Temple Lang, since it provides such an interface. We also vastly
preferred Gtk over Tcl/Tk since its widgets are more sophisticated,
which we found results in a much shorter development time.

We wanted to provide a system that was usable and extensible by
relatively naive end users and hence have provided a graphical user
interface (GUI) which they can use to both carry out an analysis, but
also supports a variety of extensions, such as changing the actions
performed in response to events. We also wanted to support a more
sophisticated developer who might want to take our basic infrastructure
and create an interactive system, say for the interactive analysis of
time series data, using our design. Hence, all commands are
available both through the GUI and through command line functions.

@@

We first give our definitions of linked views, interactive views and
linked data sets. Our model is typically a statistical data
matrix which represents data on a number of variables, for a
number of samples.  By \textit{linked view} we mean that if a
component that represents data (such as a point on a plot or a row in
a spreadsheet) changes its appearance on one view, then the
corresponding component in all other views also changes appearance.
In our example, correspondence would be measured through the sample,
but there are other possibilities, as we shall explain later.  As an
example, suppose that the model consists of the gene expression levels
from say 50 samples and 10,000 genes.  The model can be represented as
a matrix of 50 columns and ten thousand rows.  We have two choices for
linking, either the entities can be the samples, or they can be the
genes. If the entities are the genes then when two scatter plots are
created as views, where the first plot displays sample one versus sample two 
and the second displays sample three versus sample four, then the points
on each plot that referred to the same gene are corresponding
components.  If instead we had chosen to link on samples, then we
could produce pairwise plots of gene expression values and would link
them through the sample IDs.

@@

By \textit{linked data sets} we mean that the data have more than one
conceptual grouping and that there is a relationship between these
different conceptual groupings.  Linked data sets are 
familiar to users of relational databases.  Data in a relational
database are typically stored in many linked tables with rows in
different tables linked to each other through a unique identifier
called a key.  In this example, each table can be considered a data set
and they are linked through the keys.  But the idea is more general
and we shall discuss other situations where the linked data sets do
not easily correspond to tables, however, we must have a map that goes
from the entities in one data set to those in the other as well as the
inverse map, in order to propagate signals between the MVCs.

Consider Figure \ref{Fig:firstMP}, which provides a cartoon
description of the steps that occur when a user interacts with a
view. In the cartoon there are only two views, one of data set A and the
other of data set B, which is linked to data set A.  When a user
interacts with the scatter plot view of data set B, the response to
this interaction is for data set B to update itself (this is indicated
on the figure as a circled 1).  In Step 2, the scatter plot view is
told to update itself.  Next, in Step 3, because data set B is linked
to data set A, data set B sends a message to data set A.  In Step 4,
the information from data set B is converted into information that
data set A understands.  In Step 5, this information is received by
data set A and data set A updates itself.  Finally, in Step 6, data
set A tells its view to update itself.

@@

%%FIXME: please update the figure so that it has A and B, not 1 and 2
%%for the data sets - it was too confusing
%%EAW: it's changed
\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.3in, width=5in]{firstMessagePassing.jpg}
    \caption{ Linked views of linked data sets.  When a user interacts with
      the y versus x scatter plot, the first step, indicated by a circled one,
      is to tell data set B to update.  Then all views of data set B must be
      updated.  Next any data that are linked to data set B must be notified
      that data set B changed.  This information from data set B must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set A is notified that it should
      update in response to the change in data set B and then all views of
      data set A are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=0.9in, width=3.5in]{newDatabaseTables.jpg}
    \caption{ Example of Relational Database Tables. }
    \label{Fig:DBTab}
  \end{center}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=1.2in, width=1.5in]{newOneDatabaseTable.jpg}
    \caption{ Example of a Combined Table. }
    \label{Fig:OneDBTab}
  \end{center}
\end{figure}

@@

@@

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=5in, width=6in]{MultipleMVC.jpg}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

@@

\section{MVC Design Details}
\label{Sec:OneMVC}

@@

In the MVC design, the model stores the data and among other things it
supports all updating and other mechanisms that might be needed.
Information can be sent to the model to add, update, and remove data
and after every change the model is responsible for sending messages
to the views indicating which changes have occurred.  In our
implementation we created a single virtual class that represents an
abstract model and that has storage locations for data that will be
needed by all models. Then, we created a number of subclasses, one for
each type of model. In Figure~\ref{Fig:Model} we show the inheritance
structure for the three types of models we have implemented, one based
on data frames, one based on graphs and one based on the expression
set class of data used by the Bioconductor project \citep{BIOC}.
%%FIXME: we need to send the complete paper2 to the BioC tech report
%%series and then cite it here, saying something like: Complete
%%implementation details are provided in XXX.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3in, width=4.7in]{ModelClass.jpg}
    \caption{ Inheritance for Model Objects. }
    \label{Fig:Model}
  \end{center}
\end{figure}

@@

The view classes represent the visual depictions, or renderings, of
the model.  There are some data that are required by all views and all
must respond to a certain subset of events so we have implemented the view
classes by first creating a single, virtual class which the specialized
views extend.  As with our model classes, this allows us to define
methods and data storage once, for the virtual class, and only extend
it for views that specifically require additional data storage, or
specialized handling.  The object model for the view classes is shown
in Figure \ref{Fig:View}.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=4in, width=5in]{newViewClass.jpg}
    \caption{ Inheritance for View Objects. }
    \label{Fig:View}
  \end{center}
\end{figure}

As for specific view classes, the \Rclass{spreadView} class represents
a spreadsheet view of the data.  This view only makes sense for models
that are a two dimensional data structure, such as a matrix or a data
frame. Other views we have implemented include: scatter plots
(\Rclass{sPlotView}), graphs (\Rclass{graphView}), and heatmaps
(\Rclass{heatmapView}).

Since all views are expected to be interactive they have methods that
correspond to the different supported user interactions.  The events
we are most interested in responding to include key press events,
where a certain configuration of keys are pressed; button click
events, where one of the mouse buttons is clicked over the view; and
mouse movement events, where the cursor moves over the view.  When an
event occurs it creates a signal, that is caught by the signal
handler.  The signal handler then calls the callback function and it
is in the callback function that the corresponding method is called.
For example, when the user clicks the mouse button and the cursor is
over a scatter plot view, it is the scatter plot's
\Rfunction{clickEvent} method that is invoked.  Then within the
\Rfunction{clickEvent} method, the \Rfunction{identifyView} method is called
to convert the pixel coordinates into an object in the plot.

All views need an update method so they can alter themselves when the
model sends them a message that the data have been altered. The
actions taken depend on the view, for example for a spreadsheet view
the method may select or unselect a row on the spreadsheeet while for
a scatter plot view the method may change the appearance of a point on
the plot.  The inheritance structure of the view classes is
extensible.  For instance, if a user wanted to add a new plot view,
say a density plot, they can create a new view class that inherits
from \Rclass{plotView}, perhaps called \Rclass{dPlotView}.

@@

In our implementation we provide a control window, for users that want
to interact via the GUI.  It has menu items corresponding to
\texttt{File}, \texttt{Edit}, \texttt{Display}, \texttt{Event}, and
\texttt{Window}.  These menus always appear on the control window, but
the items under each menu may change depending on the type of model
that is active.  For example, the \texttt{Display} menu allows users
to create views of the active model, but only certain views are
appropriate for a type of model, as mentioned in
Section~\ref{Ssec:OneViews}.  If the active model was a data frame
model (\Rclass{dfModel}), then the views that the user could create
are a scatter plot (\Rclass{sPlotView}) and a spreadsheet
(\Rclass{spreadView}), but a graph plot (\Rclass{graphView}) does not
make sense.  Thus, when the model in the ``active'' \Robject{MVC} is
of class \Rclass{dfModel}, then the Display menu has menu items to
create a scatter plot and a spreadsheet.  Also, if the ``active''
\Robject{MVC} is of class \Rclass{dfModel}, then the command line
functions only allow the user to create a scatter plot or a spreadsheet
view of the active model, since these are the only ones we have
implemented.

@@

%%FIXME: I have tried to reword this part, but am not sure I have it
%%right, nor that I am following the design decision here.
%% EAW: When we decided that we wanted callback functions to be
%% interchangeable, we decided that the callback functions must have the same
%% signature so that calls to different callback functions would be
%% identical, as we wouldn't know ahead of time what function we would be
%% calling.  All callback functions must know which object in the plot to
%% act on, but there was no other universal information that all callback
%% functions needed.  Thus, for any other information that a callback function
%% would need, preprocessing functions were used.  As for that one parameter
%% that the callback function needed, the parameter type may be different
%% depending on the plot (and model) type - whatever is considered the object
%% of interest in a plot (and model).  Thus, for plots of data frames,
%% the object in the plot is considered a row from the data frame and for
%% plots of graphs, the object in the plot is considered a node in the graph. 
In our design a callback function takes only one parameter. Further,
within a specific MVC all callback functions must have the same
signature so that different callback functions can be interchanged to
allow different responses to an event.  The class of that parameter
depends on the type of model represented by the \Robject{MVC}.  For
example, for a graph model, the parameter corresponds to a node in the
graph, while for a data frame model, the parameter is of class
\Rclass{character}, representing a case.

%%FIXME: I thought we talked about this, and found it to be a little
%%problematic, as the way this sounds every click will ask me to set a
%%color. Somehow, some of these are essentially global options that
%%need to be set once, and then used.
%% EAW: The preprocessing functions are called once when the callback function
%% is linked to the event - they are not called each time the event occurs.
Any other information a callback function may need is set using the
preprocessing functions, which are called once when the callback function is
linked to the event.  The preprocessing functions take no
parameters.  The purpose of the preprocessing functions is to set
variables that the callback function needs.  Thus, not all callback
functions need preprocessing functions and if no preprocessing
functions are needed, then this slot is NULL.  If the callback
function sets the color of a node in a graph, then a preprocessing
function is needed to determine what color is used.  So in this
example, the preprocessing function opens a color browser to allow the
user to choose the new color.

%%FIXME: any chance you could elucidate here - what events get
%%excluded, and why are we wanting to be the "event police?"
%% EAW: Any callback function can be linked to any event, but developers may
%% only want certain callback functions available for certain events.  For
%% example, a developer may only want the color a point callback function to
%% be linked to a button click event (i.e. not the mouse over event) or the
%% tooltips callback function only linked with the mouse over event.  Thus,
%% I let the developer determine which events the callback function can
%% be linked to (if the developer wants, the callback function can be linked
%% to all events).  It is when a callback function is added as a possible
%% response to an event that the developer gets to decide which events it can
%% be a response to.  
%% Otherwise, I would just store a list of all callback functions, which could
%% be linked to any event.  I think the way it is implemented gives the
%% developer a little more control over how the callback functions are used -
%% but the developer is allowed to have his/her callback function be available
%% for all events. 
Users can link actions to events either through the GUI or via
command line functions.  The controller environment for each \Robject{MVC}
object stores a list, for each event, of appropriate callback functions that
can be linked to that event and it also stores which callback function is
currently linked to the event.  When a developer creates a new callback
function, the developer can decide which events the callback function can
potentially be linked to.  For example, the developer may only want the color
a point callback function to be available for a button click event and not a
mouse over event.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=3.2in, width=1.2in]{fullMVCClass.jpg}
    \caption{ The MVC Class. }
    \label{Fig:MVCClass}
  \end{center}
\end{figure}

%%FIXME: at this point we could perhaps even include a real example
%%with pictures of what is described below?
%% EAW: Yes, we could.  My only concern is confusing the idea of views
%% with models.  I think it's important to show the inheritance
%% structure of the models and not just views of the models.  
Thus, the \Rclass{linkedModelMVC} class not only binds the model,
view, and controller objects together, but it also shows
that the \Robject{linkedModelMVC} objects are related to each other through a
parent-child (tree) hierarchy, as shown in Figure \ref{Fig:Hier}.  In this
picture, `modelA' is the name of the original \Robject{MVC} object
(model) and from it, two \Robject{MVC} objects were created,
`modelB' and `modelC'. Then from `modelB', three
\Robject{MVC} objects were created, `modelD', `modelE', and
`modelF'.  As an example, modelA could be gene 
expression data, modelB could be gene expression data that is a subset of
modelA (based on filtering), modelC could be derived by performing
multidimensional scaling (MDS) on modelA, modelD could be the Gene Ontology
(GO) graph representing the molecular functions of the genes in modelB,
modelE could be the KEGG pathways that the genes in modelB are involved
in, and modelF could be the chromosome locations of the genes in modelB.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics[height=2.5in, width=4.2in]{Hierarchy2.jpg}
    \caption{ The Parent-Child Relationship Between \Robject{MVCs}. }
    \label{Fig:Hier}
  \end{center}
\end{figure}

@@

%%FIXME: this needs a few more words. What do these messages do, and
%%why are they the only ones we care about?
%% EAW: I've added some sentences below.
%% My concerns/questions are:
%% Obviously the other message class that could be implemented is the delete
%% message (particularly, deleting a view and deleting a model).  I remember
%% talking about this a while back and I don't recall exactly why we decided
%% not to implement them.  At the time, we discussed the delete message we did
%% not let a model be deleted (at least it wasn't implemented yet).  Message
%% classes are meant to allow communication between the components - thus,
%% deleting a view may be considered to take place solely within the
%% controller (i.e. the response to a delete event for a view is handled by
%% the contoller) and thus, a message between components is not necessary.
%% However, deleting a view seems to be the exact opposite operation as adding
%% a view and we have a message class for that operation.  
%% Plus, I'm not sure how adding a model is considered a way to communicate
%% within a MVC object since it creates the MVC object - adding a model
%% creates the MVC object that is added to the overall environment that stores
%% information about the system as a whole.
Currently, there are two types of messages within an MVC: an add
message and an update message.  Both messages request that some
modification be carried out.  Since adding a view requires different
information and methods from adding a model (and similarly with
updating views versus models), it makes sense to have two separate
message classes for these elements.  The adding a model message creates the
\Robject{MVC} object because of the one-to-one relationship between a model
and a \Robject{MVC}, whereas the adding a view message updates an
existing \Robject{MVC} object.  The updating a model message is sent by the
controller in response to an event (such as a button click event over a view)
and the updating a view message is sent by the model when the model has
changed.  Notice how these messages allow communication between components
of a \Robject{MVC} object.
